\documentclass[../../main.tex]{subfiles}

\begin{document}

\section{The density matrix, entanglement entropy and their connection to singular value decomposition (SVD)}
\label{sec:density-matrix-entanglement-svd}

\subsection{Main goal}

Prove that for $\lambda_i$ the eigenvalues of the reduced density matrix and $s_i$ the Schmidt
values,
\begin{equation}
\boxed{\,S_{\mathrm{Neum}}\equiv\Tr\bigl(\rho_A\log\rho_A\bigr)
=\sum_i\lambda_i\log\lambda_i=\sum_i s_i^2\log s_i^2\,}.
\end{equation}

\subsection{The density operator}

Let $\{p_i,\ket{\psi_i}\}$ be an ensemble of pure states, describing a system that is in one
of the states $\ket{\psi_i}$ with probability $p_i$, i.e.\ the system is not fully known. Define
the density operator as
\begin{equation}
\rho=\sum_i p_i\,\dyad{\psi_i}{\psi_i},\qquad p_i\in\mathbb R\ \&\ \sum_i p_i=1,
\end{equation}
which is a positive semi-definite operator of trace one (see for example Nielsen \& Chuang).

\subsection{Observables on a composite system}

Take a system $A$ with some observable $M$, and a system $B$, giving the composite system
$AB$ (see Box 2.6). Assume we can measure $M$ on system $A$; then a corresponding measurement
on $AB$ is $\tilde M=M\otimes\mathds{1}_B$.

\begin{proof}
Take $AB$ to be prepared in $\ket{m}\ket{\psi}$ with $M\ket{m}=m\ket{m}$. Applying the
measurement to $\ket{m}\ket{\psi}$ should yield $m$ with probability one, i.e.\ the
corresponding projector for $\tilde M$ is $P_m\otimes\mathds{1}_B$. By the spectral theorem,
$\tilde M=\sum_m m\,P_m\otimes\mathds{1}_B=M\otimes\mathds{1}_B$ for a general state of $AB$.
\end{proof}

\subsection{The reduced density matrix as a partial trace}

Measurement averages must be the same whether computed via the full $\rho^{AB}$ or via some
$\rho^{A}$ associated to system $A$, i.e.
\begin{equation}
\Trace\bigl(M\rho^A\bigr)\overset{!}{=}\Trace\bigl(\tilde M\rho^{AB}\bigr)
=\Trace\bigl((M\otimes\mathds{1}_B)\rho^{AB}\bigr).
\end{equation}
Take $\rho^A=\Trace_B(\rho^{AB})$ where
\begin{equation}
\Trace_B\bigl(\dyad{a_1}{a_2}\otimes\dyad{b_1}{b_2}\bigr)
\equiv\dyad{a_1}{a_2}\,\Trace\bigl(\dyad{b_1}{b_2}\bigr)
\qquad\&\qquad\Trace_B(\cdot)\ \text{linear.}
\end{equation}
Write $\rho^{AB}$ by using the eigenbasis of $M$ for system $A$ with $M\ket{i}=m_i\ket{i}$ and
$\braket{i}{j}=\delta_{ij}$:
\begin{align}
\rho^{AB}&=\sum_\alpha p^\alpha\dyad{\psi}{\psi}^\alpha
=\sum_\alpha p^\alpha\sum_{ijk\ell}c_{ij}^\alpha(c_{k\ell}^\alpha)^*\,
\ket{i}_A\ket{j}_B\bra{\ell}_B\bra{k}_A,\\
\rho^{A}&=\sum_\alpha\sum_{ijk\ell}p^\alpha c_{ij}^\alpha(c_{k\ell}^\alpha)^*\,
\dyad{i}{k}_A\underbrace{\Trace\bigl(\dyad{j}{\ell}_B\bigr)}_{=\delta_{j\ell}}
=\sum_{\alpha,ijk}p^\alpha c_{ij}^\alpha(c_{kj}^\alpha)^*\,\dyad{i}{k}_A.
\end{align}
Therefore
\begin{align}
\Trace\bigl(M\rho^A\bigr)
&=\sum_{\alpha,ijk}p^\alpha c_{ij}^\alpha(c_{kj}^\alpha)^*\,
\underbrace{\Trace\bigl(m_i\dyad{i}{k}_A\bigr)}_{=m_i\delta_{ik}}
=\sum_{\alpha,ij}p^\alpha m_i\,c_{ij}^\alpha(c_{ij}^\alpha)^*,
\end{align}
and similarly $\Trace\bigl((M\otimes\mathds{1}_B)\rho^{AB}\bigr)
=\sum_{\alpha,ij}p^\alpha m_i\,c_{ij}^\alpha(c_{ij}^\alpha)^*$, which proves the claim. One can
also easily show that this is the only map that conserves consistency (see Box 2.6).

\subsection{Reduced density matrix of a pure state}

Now examine the reduced density matrix for a pure state,
\begin{align}
\rho_{AB}&=\dyad{\psi}{\psi}
=\sum_{ijk\ell}c_{ij}(c_{k\ell})^*\,\ket{i}_A\ket{j}_B\bra{\ell}_B\bra{k}_A,\\
\rho_A&=\Trace_B(\rho_{AB})=\sum_{ijk}c_{ij}(c_{kj})^*\,\dyad{i}{k}_A,
\end{align}
such that the matrix representation is
\begin{equation}
\boxed{\,\rho_A=\psi\psi^\dagger\,},
\end{equation}
using that the $c_{ij}$ parametrize the matrix $\psi\in\mathbb C^{d_A\times d_B}$, i.e.\
$\ket{\psi}=\sum_{ij}c_{ij}\ket{i}_A\ket{j}_B$.

\subsection{Connection to the SVD}

Now use the SVD on $\psi$:
\begin{equation}
\psi=USV^\dagger\qquad\text{with }U\in\mathbb C^{d_A\times d_A},\ V\in\mathbb C^{d_B\times d_B}\ \text{unitary.}
\end{equation}
Then, using $V^\dagger V=\mathds 1$,
\begin{equation}
\rho_A=USV^\dagger V S U^\dagger=US^2 U^\dagger,\qquad
S^2=\diag(s_1^2,s_2^2,\dots)\in\mathbb R_{\geq 0}^{\,d_A\times d_A}.
\end{equation}
Furthermore, since $\rho_A$ is hermitian we can diagonalize it with
$\rho_A=ODO^\dagger=US^2 U^\dagger$ and $D=\diag(\lambda_1,\dots)$, so that
\begin{equation}
\boxed{\,\lambda_i=s_i^2\,},
\end{equation}
i.e.\ the eigenvalues are given by the (squared) Schmidt values.

\subsection{Functions of the density matrix via the spectral theorem}

Since $\rho_A$ is hermitian, we can use the spectral theorem to write it as
$\rho_A=\sum_i\lambda_i P_i$ where $\lambda_i$ are its eigenvalues and the eigenvectors
$\ket{u_i}$ give $P_i=\dyad{u_i}{u_i}$, since
\begin{equation}
\rho_A=UDU^\dagger=\bigl(\ket{u_1},\dots,\ket{u_n}\bigr)
\begin{pmatrix}\lambda_1& &\\ &\ddots&\\ & &\lambda_n\end{pmatrix}
\begin{pmatrix}\bra{u_1}\\\vdots\\\bra{u_n}\end{pmatrix},
\qquad\braket{u_i}{u_j}=\delta_{ij}.
\end{equation}
Thus a function $f(\cdot)$ applied to $\rho_A$ gives
\begin{align}
f(\rho_A)
&=\sum_{n=0}^{\infty}\frac{f^{(n)}(0)}{n!}\rho_A^{\,n}
\overset{(*)}{=}\sum_{n=0}^{\infty}\frac{f^{(n)}(0)}{n!}\sum_i\lambda_i^{\,n}\dyad{u_i}{u_i}\notag\\
&=\sum_i\Bigl(\sum_{n=0}^{\infty}\frac{f^{(n)}(0)}{n!}\lambda_i^{\,n}\Bigr)\dyad{u_i}{u_i}
=\sum_i f(\lambda_i)\dyad{u_i}{u_i}=U\diag\bigl(f(\lambda_i)\bigr)U^\dagger,
\end{align}
where $(*)$ uses $\rho_A^{\,n}=\sum_i\lambda_i^{\,n}\dyad{u_i}{u_i}$ (e.g.\
$\rho_A\rho_A=\sum_i\lambda_i^2\dyad{u_i}{u_i}$).

\subsection{The Von Neumann entanglement entropy}

Therefore, for the Von Neumann entanglement entropy,
\begin{align}
S=\Tr\bigl(\rho_A\log\rho_A\bigr)
&=\Tr\bigl(\rho_A\,U\log(\diag(\lambda_i))\,U^\dagger\bigr)\notag\\
&=\Tr\bigl(U\diag(\lambda_i)\underbrace{U^\dagger U}_{=\mathds 1}\log(\diag(\lambda_i))U^\dagger\bigr)\notag\\
&=\sum_i\lambda_i\log\lambda_i
=\boxed{\,\sum_i s_i^2\log s_i^2\,}.
\end{align}
$S$ is zero for two unentangled systems and non-zero for entangled systems.

\subsection{Example}

In our case, for example, $L=2$, $N_f=1$, $N_{\mathrm{ph}}=1$:
\begin{align}
&\{\ket{01},\ket{10}\}\ \&\ \{\ket{0},\ket{1}\}
\quad\text{as bases of }\mathcal H_f\ \&\ \mathcal H_{\mathrm{ph}},\\
&\Rightarrow\{\ket{01,0};\ \ket{01,1};\ \ket{10,0};\ \ket{10,1}\}
\quad\text{as basis of }\mathcal H_f\otimes\mathcal H_{\mathrm{ph}},\\
&\Rightarrow\ket{\psi}=\sum_{i=1}^{2}\sum_{j=1}^{2}\psi_{ij}\ket{i}_f\ket{j}
\quad\text{with }(\psi)_{ij}=\psi_{ij}\ \&\ \psi\in\mathbb C^{2\times2},\\
&\Rightarrow S(\rho_A)=\sum_\ell s_\ell^2\log s_\ell^2.
\end{align}
Note that since $\Trace\rho^{AB}=\Trace\rho^A=\Trace\rho^B=1$ we must have
$\sum_\ell\lambda_\ell=\sum_\ell s_\ell^2=1$.

\subsection{References}
\begin{itemize}
\item M.\ Nielsen \& I.\ Chuang, \emph{Quantum Computation and Quantum Information}
      (density operator, partial trace; Box 2.6).
\item U.\ Schollw\"ock, \emph{The density-matrix renormalization group in the age of matrix
      product states}, beginning of Chapter 4 (MPS).
\end{itemize}

\end{document}
